\input{header.tex}

\newcommand{\RR}{{\mathbb R}}
\newcommand{\bs}[1]{\hbox{\boldmath$#1$}}
\DeclareRobustCommand{\transp}{^{\rm T}}
\newcommand{\lsub}[1]{{\hbox{\footnotesize$#1$}}}
\newcommand{\lexp}[1]{{\hbox{\footnotesize$#1$}}}
\newcommand{\blankpagemark}{\centerline{\sl This page intentionally blank.}}
\DeclareRobustCommand{\prm}{^{\hbox{\footnotesize$\,\prime$}}}
\DeclareRobustCommand{\pprm}{^{\hbox{\footnotesize$\,\prime\prime$}}}
\newenvironment{m}[1]{\left[\begin{array}{@{\,}#1@{\,}}}{\end{array}\right]}
%\newcommand{\text}{\mbox}
\newcommand{\mat}[1]{\begin{bmatrix} #1 \end{bmatrix}}

%------------------------------------------------------------------------<

\begin{document}
\thispagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{edXproblem}{1805installmatlabquest}{weight=4 display_name="Matlab Practice"}

The first thing to do with Matlab is to make sure you can start it and then
we'll do some simple calculations. 

Go ahead and start Matlab. You should see a window with a 
\textbf{command prompt} (the \mlprompt{} on the left side of the window). 

\problem1 At the command prompt type the command\\
\mlcmdna{2 + 3}
Hit return and enter the result here.\\
\edXabox{expect="5" type="numerical" tolerance='0.0001'}

\problem2 Do the following computations using Matlab. 
(Enter your answers in the boxes.)\\
\mlcmdna{2*3}
\edXabox{expect="6" type="numerical" tolerance='0.0001'}

\mlcmdna{2\mlcaret 3}
\edXabox{expect="8" type="numerical" tolerance='0.0001'}

\mlcmdna{3.14\mlcaret 3.14}
\edXabox{expect="36.338" type="numerical" tolerance='0.01'}

\problem3 Use Matlab to do the following computation.
(Enter your answer in the box.)\\
\mlcmdna{100*mean(sin(30:87))}
\edXabox{expect='-2.2' type='numerical' tolerance='.1'}

\problem4 Now we'll do our first random simulation. Don't worry if you
don't fully understand the commands. We will explain them in 
the first tutorial. 

In order to simulate counting the number of heads in 1000 tosses of  a fair coin give the following sequence of commands (hit return after each one).

(In the first command the semicolon at the end 
will keep a long list of numbers from being
written to the screen. If you leave it off, you'll need to use the space bar to
page through all the numbers.)\\
\mlcmdna{x = binornd(1,.5, 1000,1);}
\mlcmdna{total = sum(x)}

You can use the up arrow to go back and run the commands again. You should get
a different total.

Give the value of the total number of heads for one of your simulations.
\edXabox{expect="500" type="numerical" tolerance='60'}

\end{edXproblem}

\end{document}




